---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# decimal

<!-- badges: start -->
[![R-CMD-check](https://github.com/pedrobtz/decimal/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/pedrobtz/decimal/actions/workflows/R-CMD-check.yaml)
[![coverage](https://raw.githubusercontent.com/pedrobtz/decimal/main/.github/badges/coverage.svg)](https://github.com/pedrobtz/decimal/actions/workflows/coverage.yml)
<!-- badges: end -->

## Overview

decimal provides exact, arbitrary-precision decimal vectors for R. If you've
ever been surprised that `0.1 + 0.2 == 0.3` is `FALSE`, this package is for
you:

```{r}
library(decimal)

0.1 + 0.2 == 0.3
decimal("0.1") + decimal("0.2") == decimal("0.3")
```

Doubles are binary fractions, so they can't represent most decimal numbers
exactly, and tiny errors accumulate as you compute. That's usually fine — but
not when you're working with money, invoices, exchange rates, or anything
else where cents have to add up. decimal uses a decimal representation and
performs arithmetic under an explicit decimal context, so any rounding is
controlled and observable.

Under the hood, decimal is built on:

* [mpdecimal](https://www.bytereef.org/mpdecimal/doc/libmpdec/), the
  battle-tested C library behind Python's `decimal` module, implementing the
  [General Decimal Arithmetic](https://speleotrove.com/decimal/decarith.pdf)
  standard.

* [vctrs](https://vctrs.r-lib.org), so decimal vectors work naturally in data
  frames, tibbles, `dplyr::mutate()`, joins, sorting, and everything else you
  already do with vectors.

Highlights:

* **Exact values.** Strings and integers are parsed exactly; promotion to a
  finer shared scale only adds trailing zeros. Values round-trip through
  `as.character()` without loss — nothing changes on the way to a CSV file or
  database column and back.

* **Full arithmetic.** `+`, `-`, `*`, `/`, `^`, `%%`, `%/%`, comparisons,
  and math functions like `abs()`, `sqrt()`, `exp()`, and `log()`, plus
  reductions `sum()`, `prod()`, `min()`, `max()`, and `mean()`.

* **Decimal-aware tools.** `quantize()` to round to a fixed number of digits
  (say, cents), `normalize()`, `fma()`, `same_quantum()`, `adjusted()`, and
  `number_class()`.

* **You control the rules.** A decimal context sets the precision, rounding
  mode, and which conditions (overflow, division by zero, …) are errors —
  see `vignette("contexts-and-signals")`.

* **Special values.** `NA`, signed zeros, infinities, and quiet and signaling
  NaNs are supported throughout.

## Installation

Install the released version from CRAN:

```{r}
#| eval: false
install.packages("decimal")
```

## Usage

Create decimal vectors from strings (exact, and the recommended way) or
integers, and use them like any other numeric vector:

```{r}
library(decimal)

x <- decimal(c("1.20", "2.30", "3.40"))
x

sum(x)
mean(x)
```

Decimal vectors are first-class citizens in tibbles and dplyr pipelines:

```{r, message = FALSE}
library(dplyr)

sales <- tibble::tibble(
  item  = c("coffee", "bagel", "juice"),
  price = decimal(c("2.50", "1.25", "3.95")),
  qty   = c(3L, 2L, 1L)
)

sales |>
  mutate(total = price * qty) |>
  summarise(revenue = sum(total))
```

Exactness matters most when small errors compound — literally, in the case
of interest:

```{r}
principal <- decimal(c("1000.00", "2500.00", "500.00"))
rate <- decimal("0.05")

balance <- principal * (1L + rate)^4L
balance

# round to cents for reporting
quantize(balance, decimal("0.01"))
```

## Learning more

* `vignette("decimal-values")` introduces decimal vectors: how to create
  them, how scale works, and the everyday operations.

* `vignette("contexts-and-signals")` covers the arithmetic context:
  precision, rounding modes, traps, and flags.

* The [General Decimal Arithmetic
  specification](https://speleotrove.com/decimal/decarith.pdf) is the
  standard that mpdecimal implements and this package follows.

## License

decimal is MIT licensed. The vendored mpdecimal library retains its own
BSD-2-Clause terms; see `inst/COPYRIGHTS`.
