---
title: "Arrow Decimal Types"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Arrow Decimal Types}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
has_arrow <- requireNamespace("arrow", quietly = TRUE) &&
  requireNamespace("tibble", quietly = TRUE)
knitr::opts_chunk$set(eval = has_arrow)
```

```{r, echo = FALSE, eval = !has_arrow, comment = NA}
message("The arrow package is not available; code output is not shown.")
```

Apache Arrow has exact decimal types — `decimal128(precision, scale)` and
`decimal256(precision, scale)` — and so does this package. They agree on what a
number *is*, so values can pass between them without losing a digit. What they
don't share is a memory layout: Arrow packs a decimal into a fixed-width
integer, while a `decimal` vector stores text and a shared scale
(`vignette("decimal-values")`).

The bridge between them is the **decimal string**. Both sides write and read
one exactly, which makes it a lossless interchange format, and the package uses
it in both directions for you. This vignette shows the crossings and where the
two type systems don't quite line up.

```{r setup}
library(decimal)
```

## A name collision to know about

`arrow` exports a `decimal()` function of its own — it builds an Arrow *type*,
not a vector — and attaching `arrow` masks this package's `decimal()`:

```r
library(arrow)
decimal(c("1.20", "2.30"))
#> Error: `precision` must be an integer
```

The masking runs whichever way you attach the two packages, so the reliable fix
is to qualify. This vignette never attaches `arrow`: every Arrow function below
is written `arrow::`, and decimal vectors are built with `decimal::decimal()`.
Adopting the same habit in a script that uses both packages will save you a
confusing error.

## From Arrow to decimal

Start with an Arrow array of exact decimals:

```{r}
a <- arrow::Array$create(
  c("100.05", "99999999999999999999.99", "0.01")
)$cast(arrow::decimal128(25, 2))
a
```

The obvious move is `as.vector()`. Don't — it converts through `double`:

```{r}
format(as.vector(a), digits = 22)
```

The first value picked up a tail of garbage, and the second lost its cents
entirely: `99999999999999999999.99` came back as `1e+20`. Twenty-two
significant digits don't fit in a double, which carries about sixteen.

`as_decimal()` takes the array directly and keeps every digit:

```{r}
as_decimal(a)
```

No `double` is involved. Arrow casts the column to text in its own exact
arithmetic, and that text is already the canonical form a `decimal` vector
stores, so the values move across without being re-parsed. That makes the exact
path about as fast as the lossy one.

### Scale comes along for free

An Arrow decimal type carries its own scale, and `as_decimal()` reads it off
the type rather than guessing from the text. That matters when a column happens
to hold only whole numbers — the declared cents survive anyway:

```{r}
whole <- arrow::Array$create(c("1", "2"))$cast(arrow::decimal128(9, 2))
as_decimal(whole)
attr(as_decimal(whole), "scale")
```

Passing `scale` overrides the type, rescaling as usual — exactly when the scale
grows, and by quantizing under the active [decimal_context()] when it shrinks:

```{r}
as_decimal(a, scale = 4)
```

### Nulls become `NA`

```{r}
as_decimal(arrow::Array$create(c("1.50", NA, "2.25"))$cast(arrow::decimal128(9, 2)))
```

### Chunked arrays work the same way

A column read from Parquet or a dataset is usually a `ChunkedArray` rather than
an `Array`. It has its own `as_decimal()` method, so no per-chunk bookkeeping is
needed:

```{r}
cs <- arrow::ChunkedArray$create(
  arrow::Array$create(c("1.25", "2.50"))$cast(arrow::decimal128(9, 2)),
  arrow::Array$create("3.75")$cast(arrow::decimal128(9, 2))
)
cs$num_chunks
as_decimal(cs)
```

## From decimal to Arrow

A `decimal` vector converts to an Arrow decimal array on its own, so it becomes
a decimal field wherever arrow infers types — `arrow::arrow_table()`,
`arrow::write_parquet()`, `arrow::write_dataset()`:

```{r}
x <- decimal::decimal(c("1.25", "2.50", "-3.75"))
arrow::as_arrow_array(x)
```

The scale is the vector's own. The precision is inferred from the values
present: the widest one here needs three digits, one before the point and two
after.

A tight precision derived from today's data may not fit tomorrow's, so for a
column you'll append to, pin a wider type:

```{r}
arrow::as_arrow_array(x, type = arrow::decimal128(20, 2))
```

Arrow refuses a cast that wouldn't fit rather than rounding silently:

```{r, error = TRUE}
arrow::as_arrow_array(x, type = arrow::decimal128(2, 2))
```

## Whole tables and Parquet files

A data frame with a decimal column becomes a table with a decimal field:

```{r}
tab <- arrow::arrow_table(
  id = 1:3,
  amount = decimal::decimal(c("100.05", "0.01", "12.30"))
)
tab$schema$GetFieldByName("amount")$type$ToString()
```

Going back, `as.data.frame()` is the wrong tool: arrow converts a decimal field
to `double`. Use `arrow_as_data_frame()`, which reads the decimal fields as
`decimal` vectors and leaves every other column to arrow:

```{r}
tibble::as_tibble(arrow_as_data_frame(tab))
```

A tibble is used here because pillar prints the column's type, which makes it
easy to confirm the decimal survived the crossing.

Parquet preserves the Arrow type, so a file written with a decimal column reads
back as one:

```{r}
path <- tempfile(fileext = ".parquet")
arrow::write_parquet(tab, path)

t2 <- arrow::read_parquet(path, as_data_frame = FALSE)
t2$schema$GetFieldByName("amount")$type$ToString()
as_decimal(t2$amount)
```

```{r, include = FALSE}
unlink(path)
```

### One wrinkle worth knowing

Arrow records an R column's attributes in the schema's metadata and reapplies
them on the way back. For a `decimal` column that means `as.data.frame()`
returns the double it would have returned anyway, now wearing the `decimal`
class — an object that is not a valid decimal vector:

```{r}
broken <- as.data.frame(tab)$amount
typeof(broken)
```

`arrow_as_data_frame()` rebuilds the column from the Arrow data and is
unaffected. Dropping the metadata with
`tab$replace_schema_metadata(NULL)` also avoids it, at the cost of every other
attribute the table was carrying.

## Where the two type systems differ

Arrow's decimals are fixed-width integers with a scale, which makes them
narrower than a `decimal` vector in two ways worth planning around.

**Infinity and NaN have no Arrow decimal.** A `decimal` vector holds them
happily; the conversion reports which element it cannot represent rather than
inventing a value:

```{r, error = TRUE}
arrow::as_arrow_array(decimal::decimal(c("1.50", "NaN")))
```

If a column can contain them, keep it as a string in Arrow, or map them to `NA`
before converting.

**Precision is capped.** `arrow::decimal128()` allows at most 38 digits and
`arrow::decimal256()` at most 76; a `decimal` vector has no such limit. The
type is chosen for you — `decimal256()` when the values need it, an error when
even that is too narrow:

```{r}
big <- decimal::decimal(
  c("12345678901234567890.12345678", "0.10000000000000000001"),
  scale = 20
)
arrow::infer_type(big)
```

```{r, error = TRUE}
arrow::infer_type(decimal::decimal(strrep("9", 90)))
```

Read back through `double`, those two wide values would have been
`12345678901234567168` and `0.10000000000000000555` — the first wrong from its
seventeenth digit, the second not 0.1 at all. Through Arrow's decimal type,
they're exact:

```{r}
as_decimal(arrow::as_arrow_array(big))
```
