---
title: "Arrow Decimal Types"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Arrow Decimal Types}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
has_arrow <- requireNamespace("arrow", quietly = TRUE) &&
  requireNamespace("tibble", quietly = TRUE)
knitr::opts_chunk$set(eval = has_arrow)
```

```{r, echo = FALSE, eval = !has_arrow, comment = NA}
message("The arrow package is not available; code output is not shown.")
```

Apache Arrow has exact decimal types — `decimal128(precision, scale)` and
`decimal256(precision, scale)` — and so does this package. They agree on what a
number *is*, so values can pass between them without losing a digit. What they
don't share is a memory layout: Arrow packs a decimal into a fixed-width
integer, while a `decimal` vector stores text and a shared scale
(`vignette("decimal-values")`).

The bridge between them is the **decimal string**. Both sides can write one and
read one exactly, which makes it a lossless interchange format. This vignette
shows how to cross in both directions, and where the two type systems don't
quite line up.

```{r setup}
library(decimal)
```

## A name collision to know about

`arrow` exports a `decimal()` function of its own — it builds an Arrow *type*,
not a vector — and attaching `arrow` masks this package's `decimal()`:

```r
library(arrow)
decimal(c("1.20", "2.30"))
#> Error: `precision` must be an integer
```

The masking runs whichever way you attach the two packages, so the reliable fix
is to qualify. This vignette never attaches `arrow`: every Arrow function below
is written `arrow::`, and decimal vectors are built with `decimal::decimal()`.
Adopting the same habit in a script that uses both packages will save you a
confusing error.

## From Arrow to decimal

Start with an Arrow array of exact decimals:

```{r}
a <- arrow::Array$create(
  c("100.05", "99999999999999999999.99", "0.01")
)$cast(arrow::decimal128(25, 2))
a
```

The obvious move is `as.vector()`. Don't — it converts through `double`:

```{r}
format(as.vector(a), digits = 22)
```

The first value picked up a tail of garbage, and the second lost its cents
entirely: `99999999999999999999.99` came back as `1e+20`. Twenty-two
significant digits don't fit in a double, which carries about sixteen.

Cast to `arrow::string()` first, then build the decimal vector:

```{r}
decimal::decimal(as.vector(a$cast(arrow::string())))
```

Every digit survives. The cast to string happens inside Arrow, in exact
arithmetic, and `decimal::decimal()` parses the text exactly — so `double`
never enters the picture.

### Scale comes along for free

An Arrow decimal type carries its own scale, and the string cast pads every
value to it. That means the resulting vector's scale already matches the source
column, with nothing for you to specify:

```{r}
a$type$scale()
attr(decimal::decimal(as.vector(a$cast(arrow::string()))), "scale")
```

If you'd rather be explicit — or you want to pin the scale even when a chunk
happens to contain only whole numbers — read it off the type and pass it along:

```{r}
decimal::decimal(
  as.vector(a$cast(arrow::string())),
  scale = a$type$scale()
)
```

### Nulls become `NA`

Arrow's nulls survive the string cast as `NA`, and land as missing decimals:

```{r}
b <- arrow::Array$create(c("1.50", NA, "2.25"))$cast(arrow::decimal128(9, 2))
decimal::decimal(as.vector(b$cast(arrow::string())))
```

### Chunked arrays work the same way

A column read from Parquet or a dataset is usually a `ChunkedArray` rather than
an `Array`. It casts the same way, and `as.vector()` flattens the chunks into
one character vector, so no per-chunk bookkeeping is needed:

```{r}
cs <- arrow::ChunkedArray$create(
  arrow::Array$create(c("1.25", "2.50"))$cast(arrow::decimal128(9, 2)),
  arrow::Array$create("3.75")$cast(arrow::decimal128(9, 2))
)
cs$num_chunks
decimal::decimal(as.vector(cs$cast(arrow::string())))
```

## From decimal to Arrow

Going the other way, write the vector out with `as.character()` and let Arrow
parse the strings into the decimal type you want:

```{r}
x <- decimal::decimal(c("1.25", "2.50", "-3.75"))
arrow::Array$create(as.character(x))$cast(arrow::decimal128(12, 2))
```

The scale to ask for is the vector's own, `attr(x, "scale")`. Precision is the
decision you have to make: it's the total number of digits, integer part
included, and Arrow refuses a cast that wouldn't fit rather than rounding
silently:

```{r, error = TRUE}
arrow::Array$create("12345.67")$cast(arrow::decimal128(4, 2))
```

Counting the digits actually present gives you the tightest type that works:

```{r}
decimal_precision <- function(x) {
  s <- as.character(x)
  s <- s[!is.na(s)]
  if (!length(s)) {
    return(1L)
  }
  s <- sub("^-", "", s)
  s <- gsub("[.]", "", s)
  max(nchar(s), 1L)
}

y <- decimal::decimal(c("1.25", "-12345.50"))
decimal_precision(y)
```

A fixed, generous precision is often the better choice for a column you'll
append to later — a tight one derived from today's data may not fit tomorrow's.

## A reusable pair

Putting both directions together:

```{r}
from_arrow_decimal <- function(x) {
  decimal::decimal(
    as.vector(x$cast(arrow::string())),
    scale = x$type$scale()
  )
}

to_arrow_decimal <- function(x, precision = decimal_precision(x)) {
  arrow::Array$create(as.character(x))$cast(
    arrow::decimal128(precision, attr(x, "scale"))
  )
}
```

They round-trip, missing values and all:

```{r}
z <- decimal::decimal(c("0.01", "999.99", NA))
identical(as.character(from_arrow_decimal(to_arrow_decimal(z))), as.character(z))
```

## Whole tables and Parquet files

The same rule governs tables: `as.data.frame()` on an Arrow table converts
decimal columns to `double`, so pull the decimal columns across yourself.

```{r}
tab <- arrow::arrow_table(
  id = arrow::Array$create(1:3),
  amount = arrow::Array$create(
    c("100.05", "0.01", "12.30")
  )$cast(arrow::decimal128(25, 2))
)

out <- tibble::tibble(
  id = as.vector(tab$id),
  amount = from_arrow_decimal(tab$amount)
)
out
```

A tibble is used here deliberately: decimal columns print through pillar, while
a base `data.frame` does not yet render them.

Parquet preserves the Arrow type, so a file written with a decimal column reads
back as one and crosses over the same way:

```{r}
path <- tempfile(fileext = ".parquet")
arrow::write_parquet(tab, path)

t2 <- arrow::read_parquet(path, as_data_frame = FALSE)
t2$schema$GetFieldByName("amount")$type$ToString()
from_arrow_decimal(t2$amount)
```

```{r, include = FALSE}
unlink(path)
```

To find the decimal columns in a schema you didn't write, check the field
types:

```{r}
types <- vapply(tab$schema$fields, function(f) f$type$ToString(), character(1))
names(tab)[grepl("^decimal", types)]
```

## Where the two type systems differ

Arrow's decimals are fixed-width integers with a scale, which makes them
narrower than a `decimal` vector in two ways worth planning around.

**Infinity and NaN have no Arrow decimal.** A `decimal` vector holds them
happily, but the cast fails rather than inventing a value:

```{r, error = TRUE}
arrow::Array$create(
  as.character(decimal::decimal(c("1.50", "NaN")))
)$cast(arrow::decimal128(9, 2))
```

If a column can contain them, either keep it as a string in Arrow or map them
to nulls before casting.

**Precision is capped.** `arrow::decimal128()` allows at most 38 digits and
`arrow::decimal256()` at most 76; a `decimal` vector has no such limit. Use
`decimal256()` for very wide values:

```{r}
big <- arrow::Array$create(
  c("12345678901234567890.12345678", "0.10000000000000000001")
)$cast(arrow::decimal256(50, 20))
from_arrow_decimal(big)
```

Read back through `double`, those two values would have been
`12345678901234567168` and `0.10000000000000000555` — the first wrong from its
seventeenth digit, the second not 0.1 at all. Through the string, they're
exact.
