---
title: "Contexts and Signals"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Contexts and Signals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(decimal)
```

`decimal` separates immutable values from the active arithmetic context.
Construction is exact; arithmetic and context-sensitive operations are where
precision, rounding, traps, and sticky flags apply.

## The default context

```{r}
get_decimal_context()
```

The defaults mirror Python's `decimal` defaults:

- precision 28
- round half even
- `Emin = -999999`
- `Emax = 999999`
- traps for invalid operation, division by zero, and overflow

## Temporary contexts

Use `with_decimal_context()` or `local_decimal_context()` to scope changes:

```{r}
ctx <- decimal_context(precision = 2L, traps = character())

with_decimal_context(ctx, decimal("1.25") + decimal("0"))
get_decimal_context()
```

The active context is restored automatically.

## Rounding and sticky flags

When traps are disabled, signals are recorded in the sticky flags:

```{r}
old <- get_decimal_context()
set_decimal_context(decimal_context(precision = 2L, traps = character()))
clear_decimal_flags()

decimal("1.25") + decimal("0")
decimal_flags()

set_decimal_context(old)
```

Sticky flags accumulate until you clear them:

```{r}
ctx <- decimal_context(precision = 2L, traps = character())
with_decimal_context(ctx, {
  clear_decimal_flags()
  decimal("1.25") + decimal("0")
  decimal("1") / decimal("8")
  decimal_flags()
})
```

## Traps

If a signal is trapped, the operation raises a classed R error:

```{r, error = TRUE}
with_decimal_context(decimal_context(traps = "division_by_zero"), {
  decimal("1") / decimal("0")
})
```

This lets calling code handle decimal failures explicitly.

## Classification and predicates

Classification is context-aware for normal versus subnormal values:

```{r}
ctx <- decimal_context(precision = 3L, emin = -2L)
with_decimal_context(ctx, {
  x <- decimal(c("1E-2", "1E-3", "-0", "NaN", "sNaN"))
  number_class(x)
})
```

Useful predicates include:

```{r}
x <- decimal(c("NaN", "sNaN", "Infinity", "-0", "1E-3"))
is_qnan(x)
is_snan(x)
is_infinite <- is.infinite(x)
is_zero(x)
is_signed(x)
```

## Notes on 0.1.0

The 0.1.0 release is intended for exact decimal workflows in ordinary R code:
data-frame columns, arithmetic, summaries, and explicit control of rounding
behavior. It does not yet expose every advanced capability from Python's
`decimal` module, but the context/signal model is already in place.
