---
title: "Contexts and Signals"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Contexts and Signals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
# Non-trapped signals warn by default (see "Reporting signals" below). Silence
# them here so the flag examples stay focused; the warning is shown on its own.
options(decimal.report_flags = FALSE)
```

When you add two decimals, how many digits should the result keep? What
happens when it doesn't fit — round it, or error? And which way should it
round? In decimal arithmetic, these are not hard-coded answers: they come from
the **context**, an explicit, session-wide setting that governs every
operation.

This vignette shows you how to inspect the context, change it (safely!), and
work with the *signals* that operations raise along the way — conditions like
`inexact`, `division_by_zero`, and `overflow`.

If you haven't yet, start with `vignette("decimal-values")`: values themselves
are immutable and construction is always exact. The context only shapes
*arithmetic* — so changing it changes how subsequent operations behave, never
the values you've already created.

```{r setup}
library(decimal)
```

## The default context

You can look at the active context at any time:

```{r}
get_decimal_context()
```

The defaults are sensible for most work:

* **precision 28** — results keep up to 28 significant digits;
* **round half even** — banker's rounding, the same default as Python;
* **`Emin = -999999` and `Emax = 999999`** — generous exponent limits;
* **traps on `invalid_operation`, `division_by_zero`, and `overflow`** — the
  conditions most likely to mean a genuine bug are errors, while routine
  rounding is not.

## Changing the context, temporarily

You'll rarely want to change the context for a whole session. Instead, scope
the change, in the same spirit as `withr::with_*()` and `withr::local_*()`.

`with_decimal_context()` applies a context to a single expression, then
restores the previous one:

```{r}
ctx <- decimal_context(precision = 2L, traps = character())

with_decimal_context(ctx, decimal("1.25") + decimal("0"))
get_decimal_context()
```

`local_decimal_context()` applies a context until the calling function
returns, which is handy when several statements need it:

```{r}
f <- function() {
  local_decimal_context(decimal_context(precision = 2L, traps = character()))
  decimal("1.234") + decimal("0")
}
f()
get_decimal_context()
```

Either way, the previous context comes back automatically when the scope ends
— even if an error interrupts it.

## Signals and sticky flags

Decimal operations *signal* noteworthy conditions: a result was rounded, a
division hit zero, a value overflowed. What happens next depends on the
context, and each signal gets one of three dispositions:

* **Trapped** — the operation raises a classed R error.
* **Reported** — the signal is recorded as a flag *and* surfaced as a warning.
  This is the default for any signal you haven't trapped.
* **Silent** — the signal is recorded as a flag only.

The flags are *sticky*: once raised, they stay raised until you clear them,
so you can run a whole computation and check afterwards what happened along
the way:

```{r}
ctx <- decimal_context(precision = 2L, traps = character())
with_decimal_context(ctx, {
  clear_decimal_flags()
  decimal("1.25") + decimal("0")
  decimal("1") / decimal("8")
  decimal_flags()
})
```

Both operations had to round to fit two significant digits, raising `inexact`
and `rounded`; the flags simply stay raised across the whole block. Use
`clear_decimal_flags()` to start a fresh slate before a computation you want
to audit.

## Reporting signals

By default, any signal that isn't trapped is also *reported* — surfaced as a
warning as it occurs, so precision loss never passes silently:

```{r}
withr::with_options(
  list(decimal.report_flags = TRUE),
  with_decimal_context(
    decimal_context(precision = 2L, traps = character()),
    decimal("1.25") + decimal("0")
  )
)
```

The warning is purely informational; the sticky flags accumulate either way.
If you'd rather check flags on your own schedule, set
`options(decimal.report_flags = FALSE)` and inspect `decimal_flags()`
directly — that's what this vignette does behind the scenes to keep the
output focused.

A few operations are exempt from the warning, because for them
`inexact`/`rounded` is the *requested* outcome, not a surprise: `quantize()`
(and `round()` and `signif()`, which build on it) exist precisely to drop
digits you named, and `sqrt()`, `exp()`, `log()`, and `log10()` are irrational
for nearly every input. Division is *not* exempt — it's only sometimes
inexact, so the warning there is genuinely informative:

```{r}
withr::with_options(
  list(decimal.report_flags = TRUE),
  with_decimal_context(decimal_context(precision = 10L, traps = character()), {
    sqrt(decimal("2"))          # inexact, but silent -- expected of sqrt()
    decimal("1") / decimal("3") # inexact, and warns -- not every division is
  })
)
```

Exempted operations still update the sticky flags as usual; only the warning
is suppressed.

## Traps: turning signals into errors

A trapped signal stops the computation with a classed R error you can handle
with `tryCatch()`:

```{r, error = TRUE}
with_decimal_context(decimal_context(traps = "division_by_zero"), {
  decimal("1") / decimal("0")
})
```

The default traps (`division_by_zero`, `invalid_operation`, `overflow`) catch
the conditions that usually indicate a bug. You can go stricter: for example,
in a context where *any* loss of precision should be an error — reconciling
ledgers, say — trap `inexact` too:

```{r, error = TRUE}
strict <- decimal_context(
  traps = c("division_by_zero", "invalid_operation", "overflow", "inexact")
)
with_decimal_context(strict, decimal("1") / decimal("3"))
```

## Classifying values

The context even shapes how values are *classified*. A finite, nonzero value
is **subnormal** when its exponent falls below the context's `emin`, meaning
fewer significant digits are available to it than `precision` allows;
otherwise it's **normal**. So classification depends on the active context,
not on the value alone — below, `1E-3` is subnormal only because `emin = -2L`
puts it out of the normal range:

```{r}
ctx <- decimal_context(precision = 3L, emin = -2L)
with_decimal_context(ctx, {
  x <- decimal(c("1E-2", "1E-3", "-0", "NaN", "sNaN"))
  number_class(x)
})
```

For quick checks there's a family of predicates:

```{r}
x <- decimal(c("NaN", "sNaN", "Infinity", "-0", "1E-3"))
is_qnan(x)
is_snan(x)
is.infinite(x)
is_zero(x)
is_signed(x)
```

`is_signed()` reports the sign bit directly, including on zero — it's how you
tell `-0` apart from `0`, since `sign()` treats both as `0`.
