---
title: "Decimal Values"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Decimal Values}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(decimal)
```

`decimal` provides exact, serializable decimal vectors backed by [`mpdecimal`](https://www.bytereef.org/mpdecimal/doc/libmpdec/).
Values are immutable; arithmetic behavior comes from the active decimal
context rather than from per-value mutable state.

## Construction and scale

Every `decimal` vector has a single shared **scale**: the number of
fractional digits stored for every element. Character and integer inputs are
exact; when `scale` is not given, it is inferred as the largest number of
fractional digits present, and every element is padded (never rounded) to
that shared scale:

```{r}
decimal(c("1.20", "-0", "Infinity", "NaN"))
decimal(c(1L, 2L, NA_integer_))
```

Because scale is a property of the vector rather than the element,
`decimal("1.2")` and `decimal("1.20")` are the same value, and combining them
produces one uniformly scaled vector:

```{r}
decimal("1.2") == decimal("1.20")
decimal(c("1.2", "1.20"))
```

Double conversion is explicit and exact, and because the exact binary value of
a double can need dozens of fractional digits, it requires an explicit or
globally configured `scale`:

```{r}
decimal("0.1")
decimal_from_double(0.1, scale = 25)
```

If you want decimal `0.1`, write `decimal("0.1")`.

For workflows that consistently use one scale, set an opt-in default. An
explicit `scale` argument still takes precedence:

```{r}
withr::with_options(
  list(decimal.default_scale = 7L),
  as_decimal(0.002)
)
```

Set it globally with `options(decimal.default_scale = 7L)`, or use
`withr::local_options(decimal.default_scale = 7L)` when the default should
apply only within the current function or test.

## Vector behavior

Decimal values are `vctrs` vectors. They combine with integers, but they do
not implicitly combine with doubles or character vectors. Combining decimal
vectors promotes to the common (largest) scale:

```{r}
vctrs::vec_c(decimal("1.20"), 2L, NA)
```

```{r, error = TRUE}
vctrs::vec_c(decimal("1.20"), 0.5)
```

## Special values

`decimal` distinguishes R missing values from decimal NaNs:

```{r}
x <- decimal(c(NA_character_, "NaN", "sNaN", "Infinity", "-0"))
x
is.na(x)
is.nan(x)
number_class(x)
```

- `NA` is an R missing value.
- `NaN` and `sNaN` are decimal NaNs.
- formatting preserves signed zero.

## Arithmetic and summaries

Arithmetic is vectorized and significance is preserved in the result:

```{r}
decimal("1.20") + decimal("2.3")
sum(decimal(c("1.20", "2.30", "3.40")))
mean(decimal(c("1", "2", "3")))
```

Some operations differ from base R because they follow decimal semantics:

```{r}
decimal("-7") %% decimal("4")
decimal("-7") %/% decimal("4")
```

The quotient truncates toward zero and the remainder has the dividend's sign.

## Decimal-specific helpers

```{r}
quantize(decimal("1.2345"), decimal("0.01"))
normalize(decimal(c("1.2300", "1.2")))
fma(decimal("2"), decimal("3"), decimal("4"))
same_quantum(decimal(c("1.20", "2.0")), decimal(c("2.30", "3.00")))
adjusted(decimal(c("1.2300", "1E-3")))
```

`same_quantum()` compares the two vectors' declared scales, since scale is a
per-vector property. `normalize()` strips shared trailing zeros down to the
finest scale the vector actually needs, without discarding any element's
significance.
