---
title: "Decimal Values"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Decimal Values}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(decimal)
```

`decimal` provides exact decimal vectors backed by
[`mpdecimal`](https://www.bytereef.org/mpdecimal/doc/libmpdec/). Values are
immutable and stored as text internally, so converting one back to a string
with `format()` or `as.character()` always reproduces it exactly -- nothing
is lost round-tripping through a CSV file or a database column. Arithmetic
behavior itself comes from the active decimal context rather than from
per-value state; see `vignette("contexts-and-signals")` for precision,
rounding, traps, and signals.

## Construction and scale

Every `decimal` vector has a single shared **scale**: the number of
fractional digits stored for every element. Character and integer inputs are
exact; when `scale` is not given, it is inferred as the largest number of
fractional digits present, and every element is padded (never rounded) to
that shared scale:

```{r}
decimal(c("1.20", "-0", "Infinity", "NaN"))
decimal(c(1L, 2L, NA_integer_))
```

Because scale is a property of the vector rather than the element,
`decimal("1.2")` and `decimal("1.20")` are the same value, and combining them
produces one uniformly scaled vector:

```{r}
decimal("1.2") == decimal("1.20")
decimal(c("1.2", "1.20"))
```

Double conversion is explicit and exact, and because the exact binary value of
a double can need dozens of fractional digits, it requires an explicit or
globally configured `scale`:

```{r}
decimal("0.1")
decimal_from_double(0.1, scale = 25)
```

If you want decimal `0.1`, write `decimal("0.1")`.

For workflows that consistently use one scale, set an opt-in default. An
explicit `scale` argument still takes precedence:

```{r}
withr::with_options(
  list(decimal.default_scale = 7L),
  as_decimal(0.002)
)
```

Set it globally with `options(decimal.default_scale = 7L)`, or use
`withr::local_options(decimal.default_scale = 7L)` when the default should
apply only within the current function or test.

## Vector behavior

Decimal values are `vctrs` vectors. They combine with integers, but they do
not implicitly combine with doubles or character vectors. Combining decimal
vectors promotes to the common (largest) scale:

```{r}
vctrs::vec_c(decimal("1.20"), 2L, NA)
```

```{r, error = TRUE}
vctrs::vec_c(decimal("1.20"), 0.5)
```

Both are rejected rather than silently coerced, because turning a double or a
string into a `decimal` isn't a lossless type promotion the way integer is:
a double needs a scale decided for it (see above), and a string needs to be
parsed, which can fail outright on invalid input. Call `as_decimal()` or
`decimal()` explicitly instead.

## Special values

`decimal` distinguishes R missing values from decimal NaNs:

```{r}
x <- decimal(c(NA_character_, "NaN", "sNaN", "Infinity", "-0"))
x
is.na(x)
is.nan(x)
number_class(x)
```

- `NA` is an R missing value, absent from the computation entirely.
- `NaN` (quiet NaN) and `sNaN` (signaling NaN) are decimal not-a-number
  values that participate in arithmetic. A quiet `NaN` propagates silently;
  an `sNaN` raises the `invalid_operation` signal, which is trapped (an
  error) under the default context -- see `vignette("contexts-and-signals")`.
- `is.na()` returns `TRUE` for both R's `NA` and decimal NaNs, matching base
  R's own `is.na(NaN)`. `is.nan()` is what distinguishes a decimal NaN from
  an R-level missing value.
- formatting preserves signed zero (`-0` prints as `-0`, not `0`).

## Arithmetic and summaries

Arithmetic is vectorized and significance is preserved in the result:

```{r}
decimal("1.20") + decimal("2.3")
sum(decimal(c("1.20", "2.30", "3.40")))
mean(decimal(c("1", "2", "3")))
```

`%%` and `%/%` differ from base R: base R's integer division floors toward
negative infinity, while decimal division truncates toward zero, per General
Decimal Arithmetic:

```{r}
-7 %% 4                        # base R: floored, remainder takes divisor's sign
decimal("-7") %% decimal("4")  # decimal: truncated, remainder takes dividend's sign

-7 %/% 4
decimal("-7") %/% decimal("4")
```

## Decimal-specific helpers

`quantize()` rounds `x` to the scale declared by `quantum`, using the active
context's rounding mode -- the same operation behind `round()` and
`signif()`:

```{r}
quantize(decimal("1.2345"), decimal("0.01"))
```

`normalize()` strips shared trailing zeros down to the finest scale the
vector actually needs, without discarding any element's significance:

```{r}
normalize(decimal(c("1.2300", "1.2")))
```

`fma()` computes `a * b + c` as one fused operation with a single rounding
at the end, rather than rounding once after the multiplication and again
after the addition:

```{r}
fma(decimal("2"), decimal("3"), decimal("4"))
```

`same_quantum()` reports whether two vectors share the same declared scale,
since scale is a per-vector rather than per-value property:

```{r}
same_quantum(decimal(c("1.20", "2.0")), decimal(c("2.30", "3.00")))
```

`adjusted()` returns each value's adjusted exponent -- the exponent it would
have in scientific notation with a single digit before the point:

```{r}
adjusted(decimal(c("1.2300", "1E-3")))
```
