---
title: "Decimal Values"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Decimal Values}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(decimal)
```

`decimal` provides exact, serializable decimal vectors backed by [`mpdecimal`](https://www.bytereef.org/mpdecimal/doc/libmpdec/).
Values are immutable; arithmetic behavior comes from the active decimal
context rather than from per-value mutable state.

## Construction

Character and integer inputs are exact:

```{r}
decimal(c("1.20", "-0", "Infinity", "NaN"))
decimal(c(1L, 2L, NA_integer_))
```

Double conversion is explicit and exact. This matters because a binary double
and a decimal literal are not the same thing:

```{r}
decimal("0.1")
as_decimal(0.1)
```

If you want decimal `0.1`, write `decimal("0.1")`.

## Vector behavior

Decimal values are `vctrs` vectors. They combine with integers, but they do
not implicitly combine with doubles or character vectors:

```{r}
vctrs::vec_c(decimal("1.20"), 2L, NA)
```

```{r, error = TRUE}
vctrs::vec_c(decimal("1.20"), 0.5)
```

## Special values

`decimal` distinguishes R missing values from decimal NaNs:

```{r}
x <- decimal(c(NA_character_, "NaN", "sNaN", "Infinity", "-0"))
x
is.na(x)
is.nan(x)
number_class(x)
```

- `NA` is an R missing value.
- `NaN` and `sNaN` are decimal NaNs.
- formatting preserves signed zero.

## Arithmetic and summaries

Arithmetic is vectorized and significance is preserved in the result:

```{r}
decimal("1.20") + decimal("2.3")
sum(decimal(c("1.20", "2.30", "3.40")))
mean(decimal(c("1", "2", "3")))
```

Some operations differ from base R because they follow decimal semantics:

```{r}
decimal("-7") %% decimal("4")
decimal("-7") %/% decimal("4")
```

The quotient truncates toward zero and the remainder has the dividend's sign.

## Decimal-specific helpers

```{r}
quantize(decimal("1.2345"), decimal("0.01"))
normalize(decimal(c("1.2300", "-0.00")))
fma(decimal("2"), decimal("3"), decimal("4"))
same_quantum(decimal("1.20"), decimal("2.30"))
adjusted(decimal(c("1.2300", "1E-3")))
```

## Compatibility boundary

Version 0.1.0 implements the core workflow: exact values, contexts,
arithmetic, comparison, summaries, and special-value handling. Some advanced
Python `decimal` features remain deferred, especially total-order APIs and the
larger toolbox of representational operations.
